---
title: "Clumping preparation meQTLs"
output: html_document
date: "2022"
---

prepare files for clumping 

# load packages
```{r}
library(here)
library(data.table)
library(dplyr)
library(tibble)
library(tidyr)
```

# load functions
```{r}
source(here("03_Code_Scripts", "functions.R"))
```

# load data

QTL analyses results - BootstrapQTL

```{r}
# meQTL

# Bootstrap QTL (data frame with results for significant hits)
load(here("02_Data/MatrixEQTL_Output", "bme_meqtl_placenta.Rdata"))
load(here("02_Data/MatrixEQTL_Output", "bme_meqtl_cvs.Rdata"))
```

# meQTLs

## placenta

### for clumping (BootstrapQTL Results)
here we have such large data that we cannot save a file for every single cpgs
```{r}
dim(bme_meqtl_placenta)
length(bme_meqtl_placenta$eSNPs < 0.05)
```
7,403,684 associations

```{r}
unique_eCpGs_placenta <- unique(bme_meqtl_placenta$eGene)
```
109,846 unique eCpGs

```{r}
write.table(unique_eCpGs_placenta, here("02_Data/clumping/placenta-meqtl_filesforclump", "unique_eCpGs_placenta.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
write.table(bme_meqtl_placenta, here("02_Data/clumping/placenta-meqtl_filesforclump", "bme_meqtl_placenta.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
splitbydupcpgs_bme_meqtl_placenta <- split(bme_meqtl_placenta, duplicated(bme_meqtl_placenta$eGene) | duplicated(bme_meqtl_placenta$eGene, fromLast = TRUE))
# splitbydupcpgs_bme_meqtl_placenta[[1]] is with only once occuring CpGs and splitbydupcpgs_bme_meqtl_placenta[[2]] with multiple
nrow(splitbydupcpgs_bme_meqtl_placenta[[1]])
nrow(splitbydupcpgs_bme_meqtl_placenta[[2]])
eCpGs_single_associations <- splitbydupcpgs_bme_meqtl_placenta[[1]]$eGene
```

7,052 CpGs have only one SNP and can stay like this
```{r}
clumped_set1_meqtl_placenta <- splitbydupcpgs_bme_meqtl_placenta[[1]]
```

```{r}
unique_eCpGs_placenta_for_clump <- setdiff(unique_eCpGs_placenta, eCpGs_single_associations) 
length(unique_eCpGs_placenta_for_clump)
```

102,794 unique CpGs where we need clump
```{r}
bme_meqtl_placenta_for_clump <- bme_meqtl_placenta[bme_meqtl_placenta$eGene %in% unique_eCpGs_placenta_for_clump, ]
length(unique(bme_meqtl_placenta_for_clump$eGene))
dim(bme_meqtl_placenta_for_clump) # 7,396,632
```

```{r}
write.table(unique_eCpGs_placenta_for_clump, here("02_Data/clumping/placenta-meqtl_filesforclump", "unique_eCpGs_placenta_for_clump.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
write.table(bme_meqtl_placenta_for_clump, here("02_Data/clumping/placenta-meqtl_filesforclump", "bme_meqtl_placenta_for_clump.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```


## cvs

#### for clumping (BootstrapQTL Results)
here we have such large data that we cannot save a file for every single cpgs
```{r}
dim(bme_meqtl_cvs)
length(bme_meqtl_cvs$eSNPs < 0.05)
```
7,403,684 associations

```{r}
unique_eCpGs_cvs <- unique(bme_meqtl_cvs$eGene)
```
57,426 unique eCpGs


```{r}
write.table(unique_eCpGs_cvs, here("02_Data/clumping/cvs-meqtl_filesforclump", "unique_eCpGs_cvs.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
write.table(bme_meqtl_cvs, here("02_Data/clumping/cvs-meqtl_filesforclump", "bme_meqtl_cvs.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
splitbydupcpgs_bme_meqtl_cvs <- split(bme_meqtl_cvs, duplicated(bme_meqtl_cvs$eGene) | duplicated(bme_meqtl_cvs$eGene, fromLast = TRUE))
# splitbydupcpgs_bme_meqtl_placenta[[1]] is with only once occuring CpGs and splitbydupcpgs_bme_meqtl_placenta[[2]] with multiple
nrow(splitbydupcpgs_bme_meqtl_cvs[[1]])
nrow(splitbydupcpgs_bme_meqtl_cvs[[2]])
eCpGs_single_associations_cvs <- splitbydupcpgs_bme_meqtl_cvs[[1]]$eGene
```

4,554 CpGs have only one SNP and can stay like this
```{r}
clumped_set1_meqtl_cvs <- splitbydupcpgs_bme_meqtl_cvs[[1]]
```

```{r}
unique_eCpGs_cvs_for_clump <- setdiff(unique_eCpGs_cvs, eCpGs_single_associations_cvs) 
length(unique_eCpGs_cvs_for_clump)
```

52872 unique CpGs where we need clump
```{r}
bme_meqtl_cvs_for_clump <- bme_meqtl_cvs[bme_meqtl_cvs$eGene %in% unique_eCpGs_cvs_for_clump, ]
length(unique(bme_meqtl_cvs_for_clump$eGene))
dim(bme_meqtl_cvs_for_clump) # 2,959,679
```

```{r}
write.table(unique_eCpGs_cvs_for_clump, here("02_Data/clumping/cvs-meqtl_filesforclump", "unique_eCpGs_cvs_for_clump.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```

```{r}
write.table(bme_meqtl_cvs_for_clump, here("02_Data/clumping/cvs-meqtl_filesforclump", "bme_meqtl_cvs_for_clump.txt"), quote = F, row.names = F, col.names=F, sep = "\t")
```


