---
title: "Basic Results eQTMs from MatrixEQTL and BootstrapQTL, common probes & samples, scaled"
author: Linda Dieckmann
date: January 2023
output: github_document
---

In this notebook we take take a look at basic results from MatrixEQTL vs. BootstrapQTL for eQTMs for the samples in both CVS + Placenta

# load packages
```{r, warning=FALSE}
library(here)
library(MatrixEQTL)
library(BootstrapQTL)
library(ggplot2)
library(dplyr)
library(cowplot)
library(scales)
library(data.table)
```

If you update package(s) call 
  renv::snapshot() 
to save the current picture of the project.

```{r}
writeLines(capture.output(sessionInfo()), here("Analyses/Scripts/sessionInfos/", "sessionInfo_Script_3_3_cp.txt"))
```

```{r}
source(here("Analyses/Scripts/", "functions.R"))
```

# load data

QTL analyses results

```{r}
# eQTM

# MatrixEQTL
load(here("02_Data/MatrixEQTL_Output", "scaled_me_eqtm_placenta_cp_samples_probes.Rdata"))
load(here("02_Data/MatrixEQTL_Output", "scaled_me_eqtm_cvs_cp_samples_probes.Rdata"))

# Bootstrap QTL (data frame with results for significant hits)
load(here("02_Data/MatrixEQTL_Output", "scaled_bme_eqtm_placenta_cp_samples_probes.Rdata"))
#load(here("02_Data/MatrixEQTL_Output", "scaled_bme_eqtm_cvs_cp_samples_probes.Rdata")) # nothing significant
```

sample names
```{r}
load(here("02_Data/prepared", "common_eqtm_samples_cvs_placenta.Rdata"))
```

gaphunter probes
```{r}
gaps_placenta <- read.table(here("02_Data/raw/gaphunter/final_data", "gaphunter_probes_placenta.txt"))
gaps_cvs <- read.table(here("02_Data/raw/gaphunter/final_data", "gaphunter_probes_cvs.txt"))
```

gene info
```{r}
load(here("02_Data/prepared/", "annotation_genes_placenta_rna.Rdata"))
load(here("02_Data/prepared/", "annotation_genes_cvs_rna.Rdata"))
```

# eQTMs
```{r}
genes_placenta <- annotation_genes_placenta_rna[,-2]
names(genes_placenta) <- c("ENSEMBL", "gene.start.position", "gene.end.position", "HGNC.symbol")

genes_cvs <- annotation_genes_cvs_rna[,-2]
names(genes_cvs) <- c("ENSEMBL", "gene.start.position", "gene.end.position", "HGNC.symbol")
```

## Placenta

### MatrixEQTL
standard eqtms output (linear model (t-statistic), only cis associations)
```{r}
scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes <- scaled_me_eqtm_placenta_cp_samples_probes$cis$eqtl
```

we add standard error and r2
```{r}
scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$se_beta <- scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$beta / scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$statistic
```

```{r}
scaled_df_eqtm_placenta_cp_samples_probes <- scaled_me_eqtm_placenta_cp_samples_probes$param$dfFull
scaled_r_eqtm_placenta_cp_samples_probes <- scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$statistic / sqrt(scaled_df_eqtm_placenta_cp_samples_probes + scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$statistic^2)
scaled_r2_eqtm_placenta_cp_samples_probes <- scaled_r_eqtm_placenta_cp_samples_probes ^ 2
```

```{r}
scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$r2 <- scaled_r2_eqtm_placenta_cp_samples_probes
```

How many tests were performed?
```{r}
scaled_me_eqtm_placenta_cp_samples_probes$cis$ntests
```

How many cis associations with FDR < 0.05 were found with MatrixEQTL?
```{r}
scaled_statistics_me_eqtm_placenta_cis_fdr5_cp_samples_probes <- scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes[scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$FDR < 0.05, ]
nrow(scaled_statistics_me_eqtm_placenta_cis_fdr5_cp_samples_probes)
```

How many unique CpGs with FDR < 0.05 were found with MatrixEQTL?
```{r}
scaled_unique_cpg_names_eqtm_placenta_fdr5_matrix_cp_samples_probes <- unique(scaled_statistics_me_eqtm_placenta_cis_fdr5_cp_samples_probes$snps)
length(scaled_unique_cpg_names_eqtm_placenta_fdr5_matrix_cp_samples_probes)
```

add a column indicating if FDR < 0.05 to the total MatrixEQTL results
```{r}
scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$FDRsig <- ifelse(scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes$FDR < 0.05, "yes", "no")
# yes if FDR < 0.05
```

How many unique genes with FDR < 0.05 were found with Matrixeqtm?
```{r}
scaled_unique_gene_names_eqtm_placenta_fdr5_matrix_cp_samples_probes <- unique(scaled_statistics_me_eqtm_placenta_cis_fdr5_cp_samples_probes$gene)
length(scaled_unique_gene_names_eqtm_placenta_fdr5_matrix_cp_samples_probes)
```


### Bootstrap QTL
How many significant cis associations were found with Bootstrap QTL (global correction p < 0.05)?
```{r}
nrow(scaled_bme_eqtm_placenta_cp_samples_probes)
```

How many unique cpgs do we have?
```{r}
scaled_unique_cpg_names_eqtm_placenta_boot_cp_samples_probes <- unique(scaled_bme_eqtm_placenta_cp_samples_probes$eSNPs)
length(scaled_unique_cpg_names_eqtm_placenta_boot_cp_samples_probes)
```

How many unqiue genes do we have?
```{r}
scaled_unique_gene_names_eqtm_placenta_boot_cp_samples_probes <- unique(scaled_bme_eqtm_placenta_cp_samples_probes$eGene)
length(scaled_unique_gene_names_eqtm_placenta_boot_cp_samples_probes)
```

```{r}
scaled_bme_eqtm_placenta_cp_samples_probes$inboot<- "yes"
# if among the bootstrapqtl results, yes
```

*combine MatrixEQTL and BootstrapQTL Data*
```{r}
scaled_eQTMs_placenta_statistics_cp_samples_probes <- merge(scaled_bme_eqtm_placenta_cp_samples_probes[,-3], scaled_statistics_me_eqtm_placenta_cis_cp_samples_probes, all = T, by.x=c("eGene", "eSNPs"), by.y=c("gene", "snps"))
names(scaled_eQTMs_placenta_statistics_cp_samples_probes)[1:2] <- c("eGene", "eCpG")
```

```{r}
scaled_eQTMs_placenta_statistics_cp_samples_probes[c("inboot")][is.na(scaled_eQTMs_placenta_statistics_cp_samples_probes[c("inboot")])] <- "no"
```

add gaphunter probes
```{r}
scaled_eQTMs_placenta_statistics_cp_samples_probes$gap_probe <- ifelse(scaled_eQTMs_placenta_statistics_cp_samples_probes$eCpG %in% gaps_placenta$V1, "yes", "no")
```

*add gene info*
```{r}
scaled_eQTMs_placenta_statistics_cp_samples_probes <- merge(scaled_eQTMs_placenta_statistics_cp_samples_probes, genes_placenta, by.x="eGene", by.y="ENSEMBL")
```

### ODDs
```{r}
total_number_tests_placenta <- scaled_me_eqtm_placenta_cp_samples_probes$cis$ntests
number_positive_placenta <- nrow(scaled_bme_eqtm_placenta_cp_samples_probes)
number_negative_placenta <- total_number_tests_placenta - number_positive_placenta

odds_placenta <- number_positive_placenta / number_negative_placenta

cat("Odds of significant association in placenta are", number_positive_placenta, "/", number_negative_placenta)
```

### export full eQTM placenta statistic table
```{r, warning=FALSE}
write.csv(scaled_eQTMs_placenta_statistics_cp_samples_probes, file = here("Analyses/Results/Descriptive", "scaled_eQTMs_placenta_statistics_cp_samples_probes.csv"), row.names = FALSE, col.names = T)
save(scaled_eQTMs_placenta_statistics_cp_samples_probes, file = here("02_Data/MatrixEQTL_Output", "scaled_eQTMs_placenta_statistics_cp_samples_probes.Rdata"))
```

### histogram plot
```{r}
scaled_hits_df_eqtms_placenta_cp_samples_probes <- data.frame(c(nrow(scaled_statistics_me_eqtm_placenta_cis_fdr5_cp_samples_probes), length(scaled_unique_cpg_names_eqtm_placenta_fdr5_matrix_cp_samples_probes), length(scaled_unique_gene_names_eqtm_placenta_fdr5_matrix_cp_samples_probes)), c(nrow(scaled_bme_eqtm_placenta_cp_samples_probes), length(scaled_unique_cpg_names_eqtm_placenta_boot_cp_samples_probes), length(scaled_unique_gene_names_eqtm_placenta_boot_cp_samples_probes)))

rownames(scaled_hits_df_eqtms_placenta_cp_samples_probes) <- c("cis eQTMs", "unique CpGs", "unique Genes")
colnames(scaled_hits_df_eqtms_placenta_cp_samples_probes) <- c("MatrixEQTL", "BootstrapQTL")

scaled_hits_df_eqtms_placenta_cp_samples_probes <- scaled_hits_df_eqtms_placenta_cp_samples_probes %>% as.matrix() %>% reshape2::melt(varnames = c("number", "method"), value.name = "cnt")
```

```{r}
barplot_counts_function(scaled_hits_df_eqtms_placenta_cp_samples_probes, "placenta", "BootstrapQTL", colors_placenta_3levels[-3], ylimits=c(0,17600)) 
```

## cvs

### MatrixEQTL
standard eqtms output (linear model (t-statistic), only cis associations)
```{r}
scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes <- scaled_me_eqtm_cvs_cp_samples_probes$cis$eqtl
```

we add standard error and r2
```{r}
scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$se_beta <- scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$beta / scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$statistic
```

```{r}
scaled_df_eqtm_cvs_cp_samples_probes <- scaled_me_eqtm_cvs_cp_samples_probes$param$dfFull
scaled_r_eqtm_cvs_cp_samples_probes <- scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$statistic / sqrt(scaled_df_eqtm_cvs_cp_samples_probes + scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$statistic^2)
scaled_r2_eqtm_cvs_cp_samples_probes <- scaled_r_eqtm_cvs_cp_samples_probes ^ 2
```

```{r}
scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$r2 <- scaled_r2_eqtm_cvs_cp_samples_probes
```

How many tests were performed?
```{r}
scaled_me_eqtm_cvs_cp_samples_probes$cis$ntests
```

How many cis associations with FDR < 0.05 were found with MatrixEQTL?
```{r}
scaled_statistics_me_eqtm_cvs_cis_fdr5_cp_samples_probes <- scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes[scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$FDR < 0.05, ]
nrow(scaled_statistics_me_eqtm_cvs_cis_fdr5_cp_samples_probes)
```

add a column indicating if FDR < 0.05 to the total MatrixEQTL results
```{r}
scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$FDRsig <- ifelse(scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes$FDR < 0.05, "yes", "no")
# yes if FDR < 0.05
```


### Bootstrap QTL
nothing significant

*combine MatrixEQTL and BootstrapQTL Data*
```{r}
scaled_eQTMs_cvs_statistics_cp_samples_probes <- scaled_statistics_me_eqtm_cvs_cis_cp_samples_probes
names(scaled_eQTMs_cvs_statistics_cp_samples_probes)[1:2] <- c("eCpG", "eGene")
```

add gaphunter probes
```{r}
scaled_eQTMs_cvs_statistics_cp_samples_probes$gap_probe <- ifelse(scaled_eQTMs_cvs_statistics_cp_samples_probes$eCpG %in% gaps_cvs$V1, "yes", "no")
```

*add gene info*
```{r}
scaled_eQTMs_cvs_statistics_cp_samples_probes <- merge(scaled_eQTMs_cvs_statistics_cp_samples_probes, genes_cvs, by.x="eGene", by.y="ENSEMBL")
```

### export full eQTM cvs statistic table
```{r}
write.csv(scaled_eQTMs_cvs_statistics_cp_samples_probes, file = here("Analyses/Results/Descriptive", "scaled_eQTMs_cvs_statistics_cp_samples_probes.csv"), row.names = FALSE)
save(scaled_eQTMs_cvs_statistics_cp_samples_probes, file = here("02_Data/MatrixEQTL_Output", "scaled_eQTMs_cvs_statistics_cp_samples_probes.Rdata"))
```

# combination CVS, Placenta
```{r}
scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs <- merge(scaled_eQTMs_placenta_statistics_cp_samples_probes, scaled_eQTMs_cvs_statistics_cp_samples_probes, by = c("eCpG", "eGene"))
cat(nrow(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs))
```

```{r}
colnames(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs) <- gsub('.x', '.placenta', colnames(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs))
colnames(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs) <- gsub('.y', '.cvs', colnames(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs))
```

### export full eQTL placenta statistic table
```{r, warning=F}
save(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs, file = here("02_Data/MatrixEQTL_Output", "scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs.Rdata"))
```

### amount associations

```{r}
number_positive_placenta / total_number_tests_placenta
#number_positive_cvs / total_number_tests_cvs
```


###################################################################################################################################################

# Placenta vs. CVS
```{r}
load(here("02_Data/MatrixEQTL_Output/", "scaled_eQTMs_placenta_statistics_cp_samples_probes.Rdata"))
load(here("02_Data/MatrixEQTL_Output", "scaled_eQTMs_cvs_statistics_cp_samples_probes.Rdata"))

load(here("02_Data/MatrixEQTL_Output", "scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs.Rdata"))
load(here("02_Data/CVS_Placenta/", "common_cpg_genes_placenta_cvs.Rdata"))
```

*common SNP-Gene pairs*
```{r}
dim(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs)
length(unique(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$eCpG))
length(unique(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$eGene))

dim(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs[scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$inboot.placenta == "yes", ])
dim(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs[scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$inboot.cvs == "yes", ])
```

```{r}
scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$abs.beta.placenta <- abs(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$beta.placenta)
scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$abs.beta.cvs <- abs(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs$beta.cvs)
```

Select those previously found 20 common eQTMs
```{r}
colnames(common_cpg_genes_placenta_cvs)[1] <- "eCpG"
common_pairs_previous_eqtms <- merge(scaled_eQTMs_statistics_cp_samples_probes_placenta_cvs, common_cpg_genes_placenta_cvs[,c("eCpG", "eGene")], by=c("eCpG", "eGene"))
nrow(common_pairs_previous_eqtms)
```
17 can be found

```{r}
# get absolute betas for placenta and cvs
common_pairs_previous_eqtms_absolute_betas <- common_pairs_previous_eqtms[ ,c("eCpG", "eGene", "abs.beta.placenta", "abs.beta.cvs")]
colnames(common_pairs_previous_eqtms_absolute_betas) <- c("eCpG", "eGene", "Placenta", "CVS")
common_pairs_previous_eqtms_absolute_betas$id_qtl <- rownames(common_pairs_previous_eqtms_absolute_betas)
```

```{r}
common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long <- reformat_long_plot(data = common_pairs_previous_eqtms_absolute_betas)
```

```{r}
PlotQQ(common_pairs_previous_eqtms_absolute_betas$Placenta-common_pairs_previous_eqtms_absolute_betas$CVS)
gghistogram(common_pairs_previous_eqtms_absolute_betas$Placenta-common_pairs_previous_eqtms_absolute_betas$CVS)
```
```{r}
common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long %>%
  group_by(tissue) %>%
  summarise_at(vars(absolute.beta), funs(mean,sd, median, IQR))
```
```{r}
t.test(common_pairs_previous_eqtms_absolute_betas$Placenta, common_pairs_previous_eqtms_absolute_betas$CVS, alternative = "greater", paired = T)
wilcox.test(common_pairs_previous_eqtms_absolute_betas$Placenta, common_pairs_previous_eqtms_absolute_betas$CVS, alternative = "greater", paired = T)
```

```{r}
# use rstatix to compute paired t test
t.test_common_eqtms <- common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long  %>% 
  t_test(absolute.beta ~ tissue, paired = TRUE) %>%
  add_significance()

# Add p-value and significance levels
t.test_common_eqtms <- t.test_common_eqtms %>% add_xy_position(x = "tissue")
```

check wilcoxon test, too
```{r}
common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long  %>% 
  wilcox_test(absolute.beta ~ tissue, paired = TRUE) %>%
  add_significance()

common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long  %>% 
  wilcox_effsize(absolute.beta ~ tissue, paired = TRUE)
```

```{r}
boxplot_paired_data(data = common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long, textlabel = "paired T-Test, t(16) = 10.98, p < 0.001, n = 17")
```
```{r}
save(common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long, file = here("02_Data/CVS_Placenta/", "common_pairs_previous_eqtms_scaled_statistics_absolute_betas_long.Rdata"))
```




